Ace: An Actively-Typed Language and Compilation Environment for High-Performance Computing

### 2013-04-24 ace/ace.tex
Updated intro

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} because they are a refinement to the concept of \emph{active libraries} (cf. Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, hindering  interoperability, the AT\&T mechanism guarantees that extensions are composable by construction.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To further demonstrate the flexibility of this mechanism, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions, including global address spaces, message passing and functional data parallelism, that can be cleanly implemented using AT\&T. First-class support for each of these has required a new language in the past. AT\&T enables compile-time logic to be coupled with user-defined types, and thus all of these first-class abstractions can safely be used within a single program or workflow.

 Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This latter mode of use supports just-in-time specialization and direct invocation of Ace functions from within Python scripts, as well as deep integration between the widely-used \verb|numpy| library (including its internal type system),  OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) and the Ace compiler itself. We describe this in Section \ref{compenv}. Achieving this deep integration makes further use of Ace's representation of types.

To demonstrate the utility of Ace for realistic scientific workloads, we briefly describe in Section \ref{casestudy} a case study that uses Ace as a compilation environment supporting a scientific simulation framework capable of orchestrating the execution of thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.

We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of some limitations of Ace at the time of writing, as well as a discussion of future work that builds upon the concept of active typing central to Ace.

>>>>

The examples in Section \ref{usage} are based on an internalization of the OpenCL kernel programming language as a library. This library is built atop a novel extensibility mechanism that we call \emph{active typechecking and translation} (AT\&T) and detail in Section \ref{att}. This mechanism relies on the core idea of representing types as metalanguage objects. We refer to such types as \emph{active types} by analogy to \emph{active libraries} \cite{activelibraries} (see Section \ref{related}). Unlike prior approaches to extensibility that take a language-global view, which impacts interoperability, the AT\&T mechanism features a guarantee that extensions are composable by construction.

%Despite its simplicity and safety, this mechanism is flexible enough to define the entirety of the OpenCL kernel language (which is derived directly from C99) and, combined with Ace's type propagation scheme, substantially enhance it's ease-of-use. 
To demonstrate the flexibility of this mechanism beyond OpenCL, we continue in Section \ref{usecases} by outlining examples of other, higher-level parallel abstractions that can be cleanly implemented using AT\&T, including global address spaces, message passing and functional data parallelism. First-class support for each of these has required a new language in the past. AT\&T allows compile-time logic to be safely included within libraries, and thus these first-class abstractions can coexist within a single program or workflow.

Ace can be used as a standalone language via the \verb|acec| compiler and also as an interactive compilation environment. This mode of use, described in Section \ref{compenv}, uses just-in-time specialization to deeply integrate the widely-adopted \verb|numpy| library (including its internal type system) and OpenCL's host API (by a mechanism also available to other host APIs, such as CUDA's) with the Ace compiler itself to enable the direct invocation of Ace functions from within Python scripts with minimal syntactic overhead. Achieving this integration also makes use of Ace's representation of types.

To demonstrate the utility of Ace for scientific workloads, we describe in Section \ref{casestudy} a case study where Ace was used as the underlying compilation environment for a scientific simulation framework. This framework was used to  specify and efficiently execute thousands of realizations of a large stochastic neurobiological circuit model on clusters of GPUs.

We conclude in Sections \ref{related} and \ref{discussion} with related work and a discussion of the limitations of Ace at the time of writing, as well as a discussion of planned future work building upon the concept of active typing.
